---
title: "MovieLens Project Report"
author: "Aileen Pan"
date: "4/11/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

The purpose of the project is to create a movie recommendation system using the MovieLens dataset. We developed algorithm using the test set and predicted movie ratings in the validation set as if they were unknown for the final test of our algorithm. Root mean squared error (RMSE) was used to evaluate how close our predictions are to the true values in the validation set. 

After spliting the train dataset further into separate training and test sets, five steps were performed to design and test the algorithm, including:
-Step 0: Explore the data
-Step 1: Predict with average ratings for all movies across all users
-Step 2: Include moviei-specific effect b_i
-Step 3: Include user-specific effect b_u
-Step 4: Regularization
-Step 5: Test final algorithm with validation set


## Methods

---------Create edx set, validation set---------

Run the code given in the Movielens project instruction.

```{r, include=FALSE}
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")

dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
                                           title = as.character(title),
                                           genres = as.character(genres))

movielens <- left_join(ratings, movies, by = "movieId")
# Validation set will be 10% of MovieLens data
set.seed(1, sample.kind="Rounding")
# if using R 3.5 or earlier, use `set.seed(1)` instead
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set
validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set
removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)
```


---------Split edx into training and test sets---------

```{r}
set.seed(755)
test_index <- createDataPartition(y = edx$rating, times = 1,
                                  p = 0.2, list = FALSE)
train_set <- edx[-test_index,]
test_set <- edx[test_index,]

#Make sure userId and movieId in test set are also in training set
test_set <- test_set %>% 
  semi_join(train_set, by = "movieId") %>%
  semi_join(train_set, by = "userId")

rm(test_index)
```

---------Set up RMSE function---------

```{r}
RMSE <- function(true_ratings, predicted_ratings){
  sqrt(mean((true_ratings - predicted_ratings)^2))
}
```

---------Setp 0: Explore the data---------

General properties of the data.
```{r,echo=FALSE}
str(edx)
```

A quick look at the data.
```{r,echo=FALSE}
head(edx)
```

The number of unique users and movies:
```{r, echo=FALSE}
edx %>% summarize(n_users = n_distinct(userId),
            n_movies = n_distinct(movieId))%>% knitr::kable()
```

TOP 10 movies that have the greatest number of ratings:
```{r, echo=FALSE}
edx %>% group_by(movieId, title) %>%
  summarize(count = n()) %>%
  arrange(desc(count)) %>%
  head(10) %>% 
  knitr::kable()
```

Distribution of the number of ratings recieved by movies:

```{r echo=FALSE}
edx%>% 
  dplyr::count(movieId) %>% 
  ggplot(aes(n)) + 
  geom_histogram(bins = 30, color = "black") + 
  scale_x_log10() + 
  ggtitle("Movies")
```

Distribution of the number of ratings given by users:

```{r,echo=FALSE}
edx %>%
  dplyr::count(userId) %>% 
  ggplot(aes(n)) + 
  geom_histogram(bins = 30, color = "black") + 
  scale_x_log10() +
  ggtitle("Users")
```


---------Step 1: Predict with average ratings for all movies across all users---------

Get the average rating.
```{r}
mu_hat <- mean(train_set$rating)
cat(mu_hat)
```

Predict all rating with the average rating and calculate RMSE.
```{r}
naive_rmse <- RMSE(test_set$rating, mu_hat)
cat(naive_rmse)
```

Create a table to store the results of RMSE.
```{r,echo=FALSE, warning=FALSE}
rmse_results <- data_frame(method = "Just the average", RMSE = naive_rmse)
rmse_results %>% knitr::kable()
```


---------Step 2: Include moviei-specific effect b_i---------

Get b_i that represents the average rating for each movie i.
```{r}
mu <- mean(train_set$rating) 
movie_avgs <- train_set %>% 
  group_by(movieId) %>% 
  summarize(b_i = mean(rating - mu))
```

Predict with fitted model.
```{r}
predicted_ratings <- mu + test_set %>% 
  left_join(movie_avgs, by='movieId') %>%
  .$b_i
```

Calculate RMSE.
```{r,echo=FALSE}
model_1_rmse <- RMSE(test_set$rating, predicted_ratings)
rmse_results <- bind_rows(rmse_results,
                          data_frame(method="Movie Effect Model",
                                     RMSE = model_1_rmse ))
rmse_results %>% knitr::kable()
```

Compared with the 'Just the average' model, the RMSE has decreased significantly. 


---------Step 3: Include user-specific effect b_u---------

Get b_u for each user.
```{r}
user_avgs <- test_set %>% 
  left_join(movie_avgs, by='movieId') %>%
  group_by(userId) %>%
  summarize(b_u = mean(rating - mu - b_i))
```

Predict with fitted model.
```{r}
predicted_ratings <- test_set %>% 
  left_join(movie_avgs, by='movieId') %>%
  left_join(user_avgs, by='userId') %>%
  mutate(pred = mu + b_i + b_u) %>%
  .$pred
```

Calculate RMSE.
```{r, echo=FALSE}
model_2_rmse <- RMSE(predicted_ratings, test_set$rating)
rmse_results <- bind_rows(rmse_results,
                          data_frame(method="Movie + User Effects Model",  
                                     RMSE = model_2_rmse ))
rmse_results %>% knitr::kable()
```

Now the RMSE continued going down after adding the user effect to the prediction. 


---------Step 4: Regularization---------

Apply a range of lambdas to find the minimum RMSE.
```{r}
lambdas <- seq(0, 10, 0.25)
rmses <- sapply(lambdas, function(l){
  mu <- mean(train_set$rating)
  b_i <- train_set %>%
    group_by(movieId) %>%
    summarize(b_i = sum(rating - mu)/(n()+l))
  b_u <- train_set %>% 
    left_join(b_i, by="movieId") %>%
    group_by(userId) %>%
    summarize(b_u = sum(rating - b_i - mu)/(n()+l))
  predicted_ratings <- 
    test_set %>% 
    left_join(b_i, by = "movieId") %>%
    left_join(b_u, by = "userId") %>%
    mutate(pred = mu + b_i + b_u) %>%
    .$pred
  return(RMSE(predicted_ratings, test_set$rating))
})
```

Plot the RMSE over lamdas.

```{r,echo=FALSE}
qplot(lambdas, rmses)
```
According to the plot, the minimum RMSE is acheived when lamda equals to 5 approximately.

Generate the lamda that has minimum RMSE.
```{r}
lambda <- lambdas[which.min(rmses)]
cat(lambda)
```

Store the result of RMSE.
```{r,echo=FALSE}
rmse_results <- bind_rows(rmse_results,
                          data_frame(method="Regularized Movie + User Effect Model",  
                                     RMSE = min(rmses)))
rmse_results %>% knitr::kable()
```

Based on the test set, the RMSE after regularize movie and user effect increased a little compared to the unregularized model.

## Results

---------Step 5: Test final algorithm with validation set---------

At this step, we will apply the regularization model to the validation set using the whole edx dataset as the training set.

Generate the new average ratings 'mu' using the whole edx dataset.
```{r}
mu_edx<- mean(edx$rating) 
cat(mu_edx)
```

Get regularized movie and user-specific effects using the 'lamda' created before.
```{r}
b_i <- edx %>%
  group_by(movieId) %>%
  summarize(b_i = sum(rating - mu_edx)/(n()+lambda))
b_u <- edx %>% 
  left_join(b_i, by="movieId") %>%
  group_by(userId) %>%
  summarize(b_u = sum(rating - b_i - mu_edx)/(n()+lambda))
```

Predict the validation set with the fitted model.
```{r}
predicted_ratings <- 
  validation %>% 
  left_join(b_i, by = "movieId") %>%
  left_join(b_u, by = "userId") %>%
  mutate(pred = mu_edx + b_i + b_u) %>%
  .$pred
```

Calculate the final RMSE.
```{r}
model_final_rmse <- RMSE(predicted_ratings, validation$rating)
cat(model_final_rmse)
```

The results of each step was presented accordingly in the method section. The RMSE of our final model is 0.8648201, which is an acceptable result. By using the whole edx dataset as the training set, we were able to further bring the RMSE to the lower, from the previous 0.8659626.


## Conclusion
The performance of the model met our requirement in terms of RMSE. However, our prediction was based on individuals and failed to account for an important source of variation related to the fact that groups of movies and groups of users have similar rating patterns. For the future work, We can observe these patterns by studying the residuals and matrix factorization.
